---
title: "tidy_coral_analysis"
author: "rpdunn"
date: "March 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r , echo=FALSE}
## libraries
library(tidyverse)
library(janitor) # install.packages('janitor')
library(skimr) # install.packages('skimr')
library(stringr) # added when we needed it for benthic data

## data filepaths/urls ----

## benthic data
benthic_url <- 'https://www.nodc.noaa.gov/archive/arc0054/0104255/1.1/data/0-data/cd08/100308OaAla03m.CSV'

## buoy data
buoy_url <- 'http://www.ndbc.noaa.gov/view_text_file.php?filename=mokh1h2010.txt.gz&dir=data/historical/stdmet/'
```

#Benthic data
This is benthic data from a series of CRAMP (Coral Reef Assessment Monitoring Program) data that includes Kaneohe Bay coral survey still images and extracted data (with larger Hawaiian Islands context):

2015. Hawaii Coral Reef Assessment and Monitoring Program (CRAMP): benthic data from digital still images made in 2008-2010 on Kauai, Oahu, Molokai, Maui, and Hawaii (NODC Accession 0104255). NOAA NCEI Oceanographic Data Archive.

```{r}
benthic_raw<-read_csv(benthic_url)
head(benthic_raw, 10)
```
This data isn't super useful. Use the janitor package to clean it up.
```{r}
benthic<-benthic_raw %>% janitor::clean_names()
names(benthic)
head(benthic)   #still have quite a few empty columns
benthic<- benthic %>% select(id_name, point, x,y,id_date)
benthic<-benthic %>% mutate(date = stringr::str_remove_all(id_date, "#"))   #could also use str_replace_all(id_date, "#","")
head(benthic)

```

```{r}
summary(benthic)
skim(benthic)
unique(benthic$id_name)
plot1<-ggplot(benthic, aes(x=date, fill=id_name)) + geom_bar()   #make a fun plot
plot1
```

#NOAA Buoy data
The buoy data come from the National Buoy Data Center. We are going to use data from the inner Kaneohe Bay buoy (station MOKH1).

```{r}
buoy_raw<-read_table2(buoy_url)
head(buoy_raw, 10)  #the first row gives units instead of data observations.
buoy<-buoy_raw    #create a backup data frame

names(buoy) <- str_c(names(buoy),                  ## current header
                  buoy %>% filter(`#YY` == "#yr"), ## row1 -- don't say buoy[1,] 
                  sep = "_")                       ## separate by `_`
## inspect
names(buoy) ## Looks a lot better

names(buoy) <- str_replace_all(names(buoy), "#", "")  # replace `#` with nothing 
names(buoy) <- str_replace_all(names(buoy), "/", "_") # replace `/` with `_`
## inspect to make sure it worked
names(buoy)
head(buoy)  #still have a redundant first row, need to remove row 1

buoy <- buoy %>%           #use filter function to remove first row, which had weird units
  filter(YY_yr != "#yr")

```
Now 'et's do some data visualization.

```{r}
ggplot(buoy, aes(WTMP_degC)) + geom_bar() + theme(axis.text.x = element_text(angle=90))

```
Look's like 999 are NA values, need to remove those.


Now let's join the benthic and buoy data.
First need to set up a date column to use to join.
```{r}
buoy<-buoy %>% 
  unite(date, c(YY_yr, MM_mo, DD_dy), sep = "-")
head(buoy)
```


```{r}
bbjoin<-left_join(benthic,buoy,by="date")
head(bbjoin)

#now let's subset this large dataset
bbjoin<-bbjoin %>% select(id_name, x,y,date,hh_hr,mm_mn,WTMP_degC)

```

```{r}
test<-filter(buoy, hh_hr =="12") %>% filter(mm_mn =="00")
head(test)

realjoin<-left_join(benthic,test, by="date")
head(realjoin)

ggplot(realjoin, aes(WTMP_degC)) + geom_bar()
```






















