---
title: "Tidy Data Tutorial"
author: "rpdunn"
date: "March 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

##Tidy data
Long-form data where variable names are at the top of columns and rows represent individual observations.
Import the Gapminder dataset and check out what the data looks like using a few different functions.

```{r echo=FALSE}
data<-read_csv("https://raw.githubusercontent.com/jules32/2017-11-30-MBARI/gh-pages/data/gapminder.csv")
  #read_csv is a tidyverse function that requires fewer arguments that read.csv (base function)
str(data)
head(data, 10)
tail(data, 5)
dim(data)
ncol(data)
nrow(data)

```

What are the values in a particular column? Use the unique() function
can select which column to look at using either $ or []

```{r , echo=FALSE}
unique(data$country)
unique(data$continent)

unique(data[,4]) #or
unique(data[,"continent"])
```



```{r}
summary(data)
data2<- subset(data, year == 2007)
summary(data2$gdpPercap)
#install.packages("skimr")    #only need to run this line if you haven't already installed this package
library(skimr)
skim(data)

```

##DPLYR
The tidy data philosophy:
Select ---> Apply ---> Combine


```{r}
filter(data,continent == "Asia")
filter(data, lifeExp >= mean(lifeExp, data=data))
data2<-select(data, -country,-pop)
filter(data2, year == 2007)
```

###the PIPE
%>%
This lets you string together multiple data manipulation functions without storing intermediate objects 
```{r}
select(data, -continent) %>% 
  filter(year == 2007)  ##when you pipe, the second function doesn't need a specific data frame to work on
```

Mutate: a way to do an operation on an entire column and then add a new column with the resulting values.
#^How is this different from vectorizing?
```{r}
data %>% mutate(gdp = gdpPercap * pop)
SA_weath<- filter(data, continent == "Americas") %>% mutate(lifetime_wealth = lifeExp*gdpPercap)
```

group_by()
Allows you to subset your dataframe using one of the columns. For instance, calculate average gdp across all 5 continents

#find max gdp for each country in the dataset
```{r}
data %>% group_by(country) %>% mutate(gdp = pop* gdpPercap, maxgdp = max(gdp)) %>%
  ungroup()
```

Summarize
```{r}
data %>% 
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarise(max_gdp = mean(gdp)) %>%
  arrange(desc(max_gdp)) %>%
  ungroup()
```

##joining 
Let's combine two data sets. In this case, adding CO2 emissions data by country for 2007.

```{r}
co2<- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv") #input co2 data
str(co2)
head(co2)
summary(co2)
```

```{r}
gap2007<- data %>% 
  filter(year == 2007) %>%    #choose only data from 2007
  full_join(co2,by="country")   #full join maintains all rows and columns, adds ots of NAs

d2007<- data %>%
  filter(year==2007)
left_join(d2007,co2,by="country")   #left join uses the gapminder 2007 data set as the "main" dataset, adds co2 data where possible
```

#tidyr
Create tidy data (long form) from non-tidy data

```{r}
 #example: the AirPassengers dataset, which is included in base R
airp<-AirPassengers
airp  #this is not in long form (tidy)
```


Use gather() to make non-tidy data into longform
```{r}
widegap<-read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv")
```

```{r}
bettergap<- widegap %>%
  gather(observation, observ_value, -continent, -country) %>%
  separate(observation, c("Variable", "Year"), sep="_", convert=T) 
#but still not there, we have multiple variable types for each year for each country. Need to make new columns for each variable type  #use spread() 
bettergap %>% 
  spread(key=Variable, value=observ_value)
```





